---
output: html_document
---
Adult Data Exploratory Visualization
=======================================================

# Data Prep

Loading necessary packages and Adult Data.

``` {r echo=TRUE, message=FALSE, warning=FALSE, packages}
# Loading useful packages
packs = c("ggplot2","ggthemes","gridExtra")
lapply(packs, function(p) {
  if (!(p %in% installed.packages()[,'Package'])) {
    install.packages(p)
  }
  else {
    gsub('package', p, 'package already installed')
  }
})
library(ggplot2)
library(ggthemes)
library(gridExtra)

# Loading adult data
cNames = c('Age','Workclass','FnlWgt','Education','EducationNum',
           'MaritalStatus','Occupation','Relationship','Race','Sex',
           'CapitalGain','CapitalLoss','HoursPerWeek','NativeCountry','Income')
adult = read.csv('../data/adult.data', header=FALSE, col.names = cNames, strip.white = TRUE)
```

# Data Overview
```{r echo=TRUE, message=FALSE, warning=FALSE, overview}
# Dimensions of data
m = nrow(adult)
n = ncol(adult)
gsub('ncol',n,gsub('nrow',m,'The dataset has nrow rows and ncol columns'))

# Feature names
colnames(adult)

# Unique classes
unique(adult$Income)

# Summary of features
summary(adult[, names(adult)!='Income'])

# Class balance
table(adult$Income) / m

# Racial balance
table(adult$Race) / m

# Gender balance
table(adult$Sex) / m

```

There are 14 features to work with, and the nature of the label "Income" shows that this task will have to be binary classification. The labels are not totally unbalanced, at 76%-24%. Racially, the dataset is quite dominated by caucasians. The dataset is also made up of about 2/3rds males.

One of the most interesting features, particularly when attempting to model Income, ought to be Education.

```{r echo=FALSE, message=FALSE, warning=FALSE, EdByGender}
# Set the ggPlot theme to something not gross
theme_set(theme_minimal(12))

# Factor levels aren't set properly; manually set
adult$Education = sapply(adult$Education, as.character)
levs = c('Preschool','1st-4th','5th-6th','7th-8th','9th','10th','11th','12th',
         'HS-grad','Some-college','Assoc-acdm','Assoc-voc','Bachelors','Masters','Prof-school','Doctorate')
adult$Education = factor(adult$Education, levels = levs)

# Education histogram
ggplot(adult, aes(Education)) +
  geom_bar(fill="#236B8E") +
  geom_vline(aes(xintercept=mean(EducationNum)), 
             color="black", linetype="dashed",size=0.75) +
  ylab("Count") + 
  ggtitle("Education Histogram w/ Mean") + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x=element_blank())

# Education split by gender
# Normalize to discount gender imbalance
edByGender = data.frame(Education=levels(adult$Education))

edByGender$NumFemale = sapply(edByGender$Education, function(e) {return(sum(adult$Education==e & adult$Sex=='Female'))})
edByGender$NumMale = sapply(edByGender$Education, function(e) {return(sum(adult$Education==e & adult$Sex=='Male'))})
edByGender$Total = with(edByGender, NumFemale+NumMale)

edByGender$EdPctOfFemale = edByGender$NumFemale / sum(edByGender$NumFemale)
edByGender$EdPctOfMale = edByGender$NumMale / sum(edByGender$NumMale)

edByGender$FemalePctOfEd = edByGender$NumFemale / edByGender$Total
edByGender$MalePctOfEd = edByGender$NumMale / edByGender$Total

edByGender$EdPctOfFemaleCuml = cumsum(edByGender$EdPctOfFemale)
edByGender$EdPctOfMaleCuml = cumsum(edByGender$EdPctOfMale)

forPlot = data.frame(Education=rep(edByGender$Education,2))
forPlot$Gender = sort(rep(c('Female','Male'), nrow(edByGender)))
forPlot$CumEdPct = c(edByGender$EdPctOfFemaleCuml, edByGender$EdPctOfMaleCuml)
forPlot$CumEdPctRev = 1 - forPlot$CumEdPct

# TODO: make cumulative line plot of Num by Gender (both up and down)
ggplot(forPlot, aes(x=rep(1:nrow(edByGender),2), y=CumEdPct, color=Gender)) + geom_line() + geom_point() +
  scale_x_continuous(labels=levels(adult$Education), breaks=1:nrow(edByGender)) +
  ylab('Prop. with at most this ed') +
  ggtitle('Cumulative Education Level By Gender') + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x=element_blank())

ggplot(forPlot, aes(x=rep(1:nrow(edByGender),2), y=CumEdPctRev, color=Gender)) + geom_line() + geom_point() +
  scale_x_continuous(labels=levels(adult$Education), breaks=1:nrow(edByGender)) +
  ylab('Prop. with at least this ed') +
  ggtitle('Cumulative Education Level By Gender') + 
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x=element_blank())
```

We see the average person graduates high school and gets "some college", and that slightly more males graduate high school, but females take the lead the rest of the way, until males catch up in grad school.

```{r echo=FALSE, message=FALSE, warning=FALSE, splitHours}
# average hours by some splitting feature
hoursSplit <- function(splittingFeature, scl) {
  groupAvgs = aggregate(adult$HoursPerWeek, by=list(adult[,splittingFeature]), mean)
  names(groupAvgs) = c('Group','Target')
  groupAvgs = groupAvgs[order(groupAvgs$Target),]
  title = gsub('group',splittingFeature, 'scl Avg Hours Per Week by group')
  if (scl) {
    groupAvgs$Target = with(groupAvgs, (Target-min(Target)) / (max(Target)-min(Target)) + 0.01)
    title = gsub('scl','Scaled',title)
  } else {
    title = gsub('scl','',title)
  }
  p = ggplot(groupAvgs, aes(x=factor(Group, levels=groupAvgs$Group), y=Target)) +
        geom_bar(stat="identity", fill='blue') +
        ggtitle(title) +
        theme(axis.text.x = element_text(angle=90, hjust=1),
              axis.title.x=element_blank())
  return(p)
}

## average hours per week for each [feature], scaled and not
# features: gender, workclass, native country, education, occupation
hoursSplit('Sex', scl=FALSE)
hoursSplit('Workclass', scl=FALSE)
hoursSplit('Workclass', scl=TRUE)
hoursSplit('NativeCountry', scl=FALSE)
hoursSplit('NativeCountry', scl=TRUE)
hoursSplit('Education', scl=FALSE)
hoursSplit('Education', scl=TRUE)
hoursSplit('Race', scl=FALSE)
hoursSplit('Race', scl=TRUE)
hoursSplit('Occupation', scl=FALSE)
hoursSplit('Occupation', scl=TRUE)
```

Unsurprisingly, the self employeed tend to put in the most hours each week. The United States and Canada are side by side in terms of hours per week. Farming/fishing is the most time consuming occupation.

```{r echo=FALSE, message=FALSE, warning=FALSE, splitIncome}
# income >=50k split on some feature
incomeSplit <- function(splittingFeature) {
  groupAvgs = aggregate(adult$Income, by=list(adult[,splittingFeature]), function(x) {return(mean(as.integer(x)-1))})
  names(groupAvgs) = c('Group','Target')
  groupAvgs = groupAvgs[order(groupAvgs$Target),]
  title = gsub('group',splittingFeature, 'Proportion >50k Income by group')
  p = ggplot(groupAvgs, aes(x=factor(Group, levels=groupAvgs$Group), y=Target)) +
        geom_bar(stat="identity", fill='blue') +
        ggtitle(title) + ylab('Pct >50k') +
        theme(axis.text.x = element_text(angle=90, hjust=1),
              axis.title.x=element_blank())
  return(p)
}

## proportion of >50k incomes for each [feature] split, not scaled
# features: gender, workclass, native country, education, occupation
incomeSplit('Sex')
incomeSplit('Workclass')
incomeSplit('NativeCountry')
incomeSplit('Education')
incomeSplit('Occupation')
```

